---
title: "EPoCH Harmonisation"
author: "Gemma Sharp"
date: "2023-01-04"
output: rmdformats::readthedown
usebookdown: true
---

# Parental smoking

Short paragraph on why we are looking at this and how we are defining it.

## Pre-conception {.tabset}

Pre-conception was defined as the period 3 months prior to pregnancy and/or the last menstrual period before pregnancy.

### Mother {.tabset .tabset-pills}

#### ALSPAC

Any relevant notes

```{r}

#CODE FOR DERIVING VARIABLES FOR PRECONCEPTION MATERNAL SMOKING IN ALSPAC

```

#### BiB

Any relevant notes

```{r}

#CODE FOR DERIVING VARIABLES FOR PRECONCEPTION MATERNAL SMOKING IN BIB

```

#### MOBA

Any relevant notes

```{r}

#CODE FOR DERIVING VARIABLES FOR PRECONCEPTION MATERNAL SMOKING IN MOBA

```

#### MCS

Any relevant notes

```{r}

#CODE FOR DERIVING VARIABLES FOR PRECONCEPTION MATERNAL SMOKING IN MCS

```

### Partner {.tabset .tabset-pills}

#### ALSPAC

Any relevant notes

```{r}

#CODE FOR DERIVING VARIABLES FOR PRECONCEPTION PARTNER SMOKING IN ALSPAC

```

#### BiB

Any relevant notes

```{r}

#CODE FOR DERIVING VARIABLES FOR PRECONCEPTION PARTNER SMOKING IN BIB

```

#### MOBA

Any relevant notes

```{r}

#CODE FOR DERIVING VARIABLES FOR PRECONCEPTION PARTNER SMOKING IN MOBA

```

#### MCS

Any relevant notes

```{r}

#CODE FOR DERIVING VARIABLES FOR PRECONCEPTION PARTNER SMOKING IN MCS

```

## First trimester {.tabset}

First trimester was defined as the first three months of pregnancy (up to 12 weeks).

### Mother {.tabset .tabset-pills}

#### ALSPAC

Any relevant notes

```{r}

#CODE FOR DERIVING VARIABLES FOR FIRST TRIMESTER MATERNAL SMOKING IN ALSPAC

```

#### BiB

Any relevant notes

```{r}

#CODE FOR DERIVING VARIABLES FOR FIRST TRIMESTER MATERNAL SMOKING IN BIB

```

#### MOBA

Any relevant notes

```{r}

#CODE FOR DERIVING VARIABLES FOR FIRST TRIMESTER MATERNAL SMOKING IN MOBA

```

#### MCS

Any relevant notes

```{r}

#CODE FOR DERIVING VARIABLES FOR FIRST TRIMESTER MATERNAL SMOKING IN MCS

```

### Partner {.tabset .tabset-pills}

#### ALSPAC

Any relevant notes

```{r}

#CODE FOR DERIVING VARIABLES FOR FIRST TRIMESTER PARTNER SMOKING IN ALSPAC

```

#### BiB

Any relevant notes

```{r}

#CODE FOR DERIVING VARIABLES FOR FIRST TRIMESTER PARTNER SMOKING IN BIB

```

#### MOBA

Any relevant notes

```{r}

#CODE FOR DERIVING VARIABLES FOR FIRST TRIMESTER PARTNER SMOKING IN MOBA

```

#### MCS

Any relevant notes

```{r}

#CODE FOR DERIVING VARIABLES FOR FIRST TRIMESTER PARTNER SMOKING IN MCS

```


## Second trimester

## Third trimester

## Any time in pregnancy

## Post-natal

# Parental Alcohol

# Parental Caffeine
