---
title: "Results"
output: word_document
---


```{r setup, include=FALSE,warning=FALSE,message=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#load data

##first questionnaires
bib_sa <- readRDS("/Volumes/MRC-IEU-research/projects/ieu2/p5/015/working/data/bib/bib_sa_pheno.rds")
bib_wb <- readRDS("/Volumes/MRC-IEU-research/projects/ieu2/p5/015/working/data/bib/bib_wb_pheno.rds")
alspac <- readRDS("/Volumes/MRC-IEU-research/projects/ieu2/p5/015/working/data/alspac/alspac_pheno.rds")
mcs <- readRDS("/Volumes/MRC-IEU-research/projects/ieu2/p5/015/working/data/mcs/mcs_pheno.rds")

library(dplyr)
library(epitools)

#select only unique pregnancies that feature biological mums with prenatal health behaviour data
bib_sa_unique <- bib_sa[!duplicated(bib_sa$PregnancyID),]
bib_sa_unique <- bib_sa_unique[is.na(bib_sa_unique$MotherID)==FALSE,]
bib_wb_unique <- bib_wb[!duplicated(bib_wb$PregnancyID),]
bib_wb_unique <- bib_wb_unique[is.na(bib_wb_unique$MotherID)==FALSE,]
alspac_unique <- data.frame(alspac[!duplicated(alspac$aln),])
mcs_unique <- data.frame(mcs[!duplicated(mcs$mcsid),])

#set all paternal data to missing if partner is not the biological dad, or if paternity status is missing or mother is unsure
alspac_unique[alspac_unique$covs_biological_father%in%c("not_bio_dad","unsure","missing"),grep(colnames(alspac_unique),pattern="father")]<-NA
mcs_unique[mcs_unique$covs_biological_father%in%c("not_bio_dad","unsure","missing"),grep(colnames(mcs_unique),pattern="father")]<-NA

# set paternal participation to 0 if mum has no partner, partner is not the biological dad (obviously there is a biological father somewhere, but he's not participating in the study)
alspac_unique[alspac_unique$covs_biological_father%in%c("not_bio_dad","no_partner"),grep(colnames(alspac_unique),pattern="paternal_participation")]<-0
mcs_unique[mcs_unique$covs_biological_father%in%c("not_bio_dad","no_partner"),grep(colnames(mcs_unique),pattern="paternal_partcipation")]<-0

# set paternal participation to missing if paternity status is missing or unclear
alspac_unique[alspac_unique$covs_biological_father%in%c("missing","unsure"),grep(colnames(alspac_unique),pattern="paternal_participation")]<-NA
mcs_unique[mcs_unique$covs_biological_father%in%c("missing","unsure"),grep(colnames(mcs_unique),pattern="paternal_partcipation")]<-NA

```

# Low paternal cohort participation

```{r , echo=FALSE,include=FALSE,warning=FALSE,message=FALSE}


df <- data.frame(cohort=c("BiB South Asian","BiB White British","ALSPAC","MCS"),mother=c(sum(is.na(bib_sa_unique$MotherID)==FALSE),sum(is.na(bib_wb_unique$MotherID)==FALSE),sum(is.na(alspac_unique$aln)==FALSE),sum(is.na(mcs_unique$mcsid)==FALSE)), father=c(sum(is.na(bib_sa_unique$FatherID)==FALSE,na.rm=T),sum(is.na(bib_wb_unique$FatherID)==FALSE,na.rm=T),sum(alspac_unique$paternal_participation,na.rm=T),sum(mcs_unique$paternal_participation,na.rm=T)))

df$percent <- (df$father/df$mother)*100


#devtools::install_github('cttobin/ggthemr')
#devtools::install_github("brooke-watson/bplots")
require(bplots)
require(ggthemr)
require(ggplot2)
require(extrafont)

ggthemr("light")

plot1<-ggplot(df) +
  geom_rect(aes(xmin=0,xmax=percent,ymin=0,ymax=1,fill="#62bba5"))+
  xlab("Percentage of cohort participating fathers")+
  ggtitle("Paternal participation in ALSPAC, BiB and MCS")+
  xlim(0,100)+
  facet_grid(cohort~., switch = "y")+
  theme_grey(base_family="Avenir")+
  theme(strip.text.y = element_text(angle = 180),axis.text.y=element_blank(),axis.ticks.y=element_blank(),panel.grid.minor = element_blank(),panel.grid.major = element_blank(),panel.spacing = unit(.7, "lines"),legend.title=element_blank(),legend.position="none",plot.title = element_text(hjust = 0.5))

```
Biological fathers participate in ALSPAC and MCS for the majority of pregnancies. Of 14,472 pregnancies in ALSPAC, and 18,497 pregnancies in MCS, 11,037 (76%) and 13,145 (71%) had a cohort participating biological father, respectively. However paternal participation was much lower in BiB: of 11,395 pregnancies where the mother had returned the baseline questionnaire, only 3131 (27%) of fathers did the same (35% White British, 22% South Asian).

```{r, echo=FALSE,include=TRUE,fig.height=4,fig.width=5,warning=FALSE,message=FALSE}
plot1
```

# Less detailed paternal data

In all three cohorts, data on paternal health behaviours is less detailed and collected less frequently than data on maternal health behaviours. (make a table in word). This makes it particularly difficult to harmonise data between different studies, and to harmonise maternal and paternal data.

# Measurement error in paternal data reported by the mother

In ALSPAC, two of our paternal variables of interest (smoking at the start of pregnancy and drinking alcohol at any point during pregnancy) were reported by both the study father himself and the study mother about the study father. 

```{r , echo=FALSE,include=TRUE,fig.height=4,fig.width=6,warning=FALSE,message=FALSE}

require(knitr)
require(stringr)
mat_report_vars <- names(alspac)[grep(names(alspac_unique),pattern="mreport")][-1]
pat_report_vars <- str_remove(mat_report_vars,"_mreport")
```

For drinking alcohol during pregnancy, of 9963 pregnancies where both the mother and father reported paternal data, both sources agreed in 98% of pregnancies (n=). 
```{r , echo=FALSE,include=TRUE,fig.height=4,fig.width=6,warning=FALSE,message=FALSE}
df <- na.omit(data.frame(self_report=alspac_unique$alcohol_father_ever_pregnancy_binary,mother_report=alspac_unique$alcohol_father_ever_pregnancy_binary_mreport))

table(df$self)
table(df$mother)
prop.table(table(df$self))
prop.table(table(df$mother))

prop.table(table(df$self==df$mother))*100

library(irr)
kappa2(apply(df,2,as.factor)) # kappa = 0.659, p=0
```

For smoking at the start of pregnancy, the overall error rate was 4.8%, and this was highest in the exposed group: 8.2% (n=284) of fathers who said they smoked at the start of pregnancy were reported as not smoking by the study mother, and 2.7% (n=160) of fathers who said they did not smoke were reported as smoking by the study mother. We were also able to explore smoking at the start of pregnancy as an ordinal variable, which showed lower rates of concordance between maternal and paternal reports at higher levels of exposure. For 67% of pregnancies where the maternal and paternal report of paternal smoking differed, the study mother reported a lower level of paternal smoking than the study father; i.e. the study mother tended to under-report paternal smoking.
```{r , echo=FALSE,include=TRUE,fig.height=4,fig.width=6,warning=FALSE,message=FALSE}
df <- na.omit(data.frame(self_report=alspac_unique$smoking_father_startpregnancy_binary,mother_report=alspac_unique$smoking_father_startpregnancy_binary_mreport))

table(df$self)
table(df$mother)
prop.table(table(df$self))
prop.table(table(df$mother))

prop.table(table(df$self==df$mother))*100

kappa2(apply(df,2,as.factor)) # kappa = 0.897, p=0

df<- na.omit(data.frame(mother_report=alspac_unique[,mat_report_vars[2]],self_report=alspac_unique[,pat_report_vars[2]]))

ggthemr("light")

ggplot(df,aes(x=self_report))+
  geom_bar(aes(fill=as.character(mother_report)),position="fill")+
  labs(y="Proportion",
       x="Self-report of paternal smoking",
       title ="Proportion of paternal smoking:\na comparison of self- and maternal-reported data in ALSPAC",
      fill="Maternal report of paternal smoking:",
      bottom = title)+
  theme_grey(base_family="Avenir")+
  theme(panel.grid.minor = element_blank(),panel.grid.major = element_blank(),legend.position="bottom", plot.title = element_text(hjust = 0.5))

prop.table(table(as.character(df$self_report)==as.character(df$mother_report)))
df_disagreement <- df[df$self_report!=df$mother_report,]
prop.table(table(df_disagreement$mother_report<df_disagreement$self_report))

kappa2(df) # kappa = 0.767, p=0 (UNWEIGHTED)
kappa2(df,"equal") # kappa = 0.794, p=0 (WEIGHTED)
kappa2(df,"squared") # kappa = 0.829, p=0 (WEIGHTED)

```

**Mat report vs pat report smoking on BW and GA**
```{r , echo=FALSE,include=TRUE,warning=FALSE,message=FALSE}
ggthemr("light")

## ALSPAC
all_res <- rbind.data.frame(c(summary(lm(scale(anthro_birthweight)~smoking_father_startpregnancy_binary,data=alspac_unique))$coef[2,],"Self report","Complete sample","ALSPAC","Birthweight","Paternal smoking at the start of pregnancy"),
      							c(summary(lm(scale(anthro_birthweight)~smoking_father_startpregnancy_binary_mreport,data=alspac_unique))$coef[2,],"Mother report","Complete sample","ALSPAC","Birthweight","Paternal smoking at the start of pregnancy"),
      							c(summary(lm(scale(scale(covs_gestage))~smoking_father_startpregnancy_binary,data=alspac_unique))$coef[2,],"Self report","Complete sample","ALSPAC","Gestational age","Paternal smoking at the start of pregnancy"),
      							c(summary(lm(scale(scale(covs_gestage))~smoking_father_startpregnancy_binary_mreport,data=alspac_unique))$coef[2,],"Mother report","Complete sample","ALSPAC","Gestational age","Paternal smoking at the start of pregnancy"),
      							c(summary(lm(scale(anthro_birthweight)~alcohol_father_ever_pregnancy_binary,data=alspac_unique))$coef[2,],"Self report","Complete sample","ALSPAC","Birthweight","Paternal alcohol in pregnancy"),
      							c(summary(lm(scale(anthro_birthweight)~alcohol_father_ever_pregnancy_binary_mreport,data=alspac_unique))$coef[2,],"Mother report","Complete sample","ALSPAC","Birthweight","Paternal alcohol in pregnancy"),
      							c(summary(lm(scale(scale(covs_gestage))~alcohol_father_ever_pregnancy_binary,data=alspac_unique))$coef[2,],"Self report","Complete sample","ALSPAC","Gestational age","Paternal alcohol in pregnancy"),
      							c(summary(lm(scale(scale(covs_gestage))~alcohol_father_ever_pregnancy_binary_mreport,data=alspac_unique))$coef[2,],"Mother report","Complete sample","ALSPAC","Gestational age","Paternal alcohol in pregnancy"))
colnames(all_res)<-c("estimate","se","t","p","parent","model","cohort","outcome","exposure")

all_res[,c("estimate","se","t","p")]<-apply(all_res[,c("estimate","se","t","p")],2,function(x) as.numeric(as.character(x)))
all_res$model <- as.character(all_res$model)
all_res$model <- factor(all_res$model, ordered=T, levels = rev(unique(all_res$model)))
all_res$parent <- as.character(all_res$parent)
all_res$parent <- factor(all_res$parent, ordered=T, levels = rev(unique(all_res$parent)))

mreport_plot<-ggplot(data=all_res, aes(x=estimate, y=1,colour=as.character(parent))) + 
 geom_vline(aes(xintercept = 0), size = .25, linetype = "dashed", colour= "grey40" ) +
  geom_errorbarh(aes(xmax = estimate+(se*1.96), xmin = estimate-(se*1.96)), height = .1, position=position_dodgev(.5)) +
  geom_point(size=3, fill="white",position=position_dodgev(.5),shape=15)+
  theme_grey(base_family = "Avenir")+
  facet_grid(exposure~outcome,switch="y")+
  labs( y = "", x = "Difference in SD outcome in offspring of mothers who smoked\nvs those who did not smoke during pregnancy") +
 theme(strip.text.y = element_text(angle=180), panel.grid.minor = element_blank(),panel.grid.major = element_blank(),panel.spacing = unit(.7, "lines"),panel.spacing.y = unit(0.1, "lines"),legend.title=element_blank(),legend.position="bottom",plot.title = element_text(hjust = 0.5),axis.text.y=element_blank(),axis.ticks.y=element_blank()) +
  labs(title= "Estimates of association between paternal health behaviours\nduring pregnancy and offspring birthweight and gestational age") 
mreport_plot
```


# Selection bias in paternal data
Mothers and babies whose partners/fathers participated in the cohorts differ from mothers and babies whose partners/fathers did not participate on several key characteristics. (obviously reword this terrible sentence!) And this was different in different cohorts.

**paternal participation ~ maternal and birth characteristics**


```{r , echo=FALSE,include=TRUE,warning=FALSE,message=FALSE,fig.height=15,fig.width=4}

# comparing proportions
calc_props <- function(df,variable){
d00 <- sum(na.omit(df$paternal_participation[which(df[,variable]==0)])==0)
d10 <- sum(na.omit(df$paternal_participation[which(df[,variable]==0)])==1)
d01 <- sum(na.omit(df$paternal_participation[which(df[,variable]==1)])==0)
d11 <- sum(na.omit(df$paternal_participation[which(df[,variable]==1)])==1)
res<-data.frame(variable=variable,p0v0=d00,p0v1=d01,p1v0=d10,p1v1=d11,percentv1_pp0=NA,percentv1_pp1=NA)
res$percentv1_pp0 <- (res$p0v1/(res$p0v1+res$p0v0))*100
res$percentv1_pp1 <- (res$p1v1/(res$p1v1+res$p1v0))*100
res$difference <- res$percentv1_pp1-res$percentv1_pp0
ORres <- summary(glm(df$paternal_participation ~ df[,variable],family="binomial"))$coefficients[2,]
res$est <- ORres[1]
res$or <- exp(ORres[1])
res$or_lower <- exp(ORres[1] - (1.96*ORres[2]))
res$or_upper <- exp(ORres[1] + (1.96*ORres[2]))
res
}

gen_new_vars <- function(df){
df$covs_mother_edu_high <- df$covs_edu_mother
df$covs_mother_edu_high[df$covs_edu_mother==3]<-1
df$covs_mother_edu_high[df$covs_edu_mother<=2]<-0

df$covs_mother_edu_low <- df$covs_edu_mother
df$covs_mother_edu_low[df$covs_edu_mother==0]<-1
df$covs_mother_edu_low[df$covs_edu_mother>=1]<-0

df$ethnicity_white <- NA
df$ethnicity_white[df$covs_ethnicity_mother_binary=="white"]<-1
df$ethnicity_white[df$covs_ethnicity_mother_binary=="not white"]<-0

if("covs_age_mother_conception" %in% names(df)){
df$maternal_age_high <-NA
df$maternal_age_high[df$covs_age_mother_conception>=35]<-1
df$maternal_age_high[df$covs_age_mother_conception<35]<-0

df$maternal_age_low <-NA
df$maternal_age_low[df$covs_age_mother_conception<=21]<-1
df$maternal_age_low[df$covs_age_mother_conception>21]<-0
}else{
  df$maternal_age_high <-NA
df$maternal_age_high[df$covs_age_mother_delivery>=35]<-1
df$maternal_age_high[df$covs_age_mother_delivery<35]<-0

df$maternal_age_low <-NA
df$maternal_age_low[df$covs_age_mother_delivery<=21]<-1
df$maternal_age_low[df$covs_age_mother_delivery>21]<-0
}

df
}

bib_sa_unique <- gen_new_vars(bib_sa_unique)
bib_wb_unique <- gen_new_vars(bib_wb_unique)
alspac_unique <- gen_new_vars(alspac_unique)
mcs_unique <- gen_new_vars(mcs_unique)

variables <- c("maternal_age_low","maternal_age_high","smoking_mother_ever_life_binary","smoking_mother_ever_pregnancy_binary","alcohol_mother_ever_pregnancy_binary","covs_mother_edu_high","covs_mother_edu_low","covs_partner_lives_with_mother_prenatal","covs_married_mother_binary","anthro_birthweight_low_binary","anthro_birthweight_high_binary","covs_preterm_binary")

df_alspac <- bind_rows(lapply(variables,calc_props,df=alspac_unique))
df_sa_bib <- bind_rows(lapply(variables,calc_props,df=bib_sa_unique))
df_wb_bib <- bind_rows(lapply(variables,calc_props,df=bib_wb_unique))
df_mcs <- bind_rows(lapply(variables,calc_props,df=mcs_unique))


df <- data.frame(varname=rep(c("mother is age <=21","mother is age >=35","mother has ever been a smoker","mother smoked in pregnancy","mother drank alcohol in pregnancy","mother has high level of education","mother has low level of education","mother lived with partner before birth","mother is married","baby's birthweight is <2.5kg","baby's birthweight is >4.5kg","baby was delivered preterm"),4),
                 pp0=c(df_alspac[,6],df_sa_bib[,6],df_wb_bib[,6],df_mcs[,6]),
                 pp1=c(df_alspac[,7],df_sa_bib[,7],df_wb_bib[,7],df_mcs[,7]),
                 difference=c(df_alspac[,8],df_sa_bib[,8],df_wb_bib[,8],df_mcs[,8]),cohort=rep(c("ALSPAC","BiB South Asian","BiB White British","MCS"),each=12),
                 same_direction1=apply(data.frame(sign(df_alspac$difference)==1,sign(df_sa_bib$difference)==1,sign(df_wb_bib$difference)==1,sign(df_mcs$difference)==1),1,sum),
                 average_for_ordering=apply(data.frame(df_alspac$difference,df_sa_bib$difference,df_wb_bib$difference,df_mcs$difference),1,mean),
                 or=c(df_alspac$or,df_sa_bib$or,df_wb_bib$or,df_mcs$or),
                 or_lower=c(df_alspac$or_lower,df_sa_bib$or_lower,df_wb_bib$or_lower,df_mcs$or_lower),
                 or_upper=c(df_alspac$or_upper,df_sa_bib$or_upper,df_wb_bib$or_upper,df_mcs$or_upper),
                 est=c(df_alspac$est,df_sa_bib$est,df_wb_bib$est,df_mcs$est)
                 )

df$same_direction2<-0
df$same_direction2[df$same_direction1%in%c(0,3)]<-1

#df <- df[order(df$average_for_ordering,df$same_direction1,df$same_direction2),]
df <- df[order(df$average_for_ordering,df$same_direction1,df$same_direction2),]
#df <- df[order(df$same_direction2,decreasing=T),]
df$varname <- factor(df$varname,ordered=T,levels=unique(df$varname))

df$label_position <- NA
df$label_position[df$or>1] <- df$or[df$or>1]+(df[df$or>1,]$or*0.05)
df$label_position[df$or<1] <- df$or[df$or<1]-(df[df$or<1,]$or*0.05)

library(ggthemr)
ggthemr("light")

  ggplot(df)+
   # geom_rect(aes(xmin=-pp0,xmax=pp1,ymin=as.numeric(as.factor(cohort))-.5,ymax=as.numeric(as.factor(cohort))+.5,fill=cohort))+
 #   geom_point(aes(x=or,y=as.numeric(as.factor(cohort)),colour=cohort))+
  geom_rect(aes(xmin=1,xmax=or,ymin=as.numeric(as.factor(cohort))-.5,ymax=as.numeric(as.factor(cohort))+.5,fill=cohort))+
    geom_text(aes(label=round(or,2),x=label_position,y=as.numeric(as.factor(cohort))),size=2.5)+
 #       geom_errorbarh(aes(xmin=or_lower,xmax=or_upper,y=as.numeric(as.factor(cohort)),colour=cohort))+
 #   facet_grid(.~varname, switch = "y",scales="free")+
 #   geom_vline(xintercept = 0,colour="grey40")+
        geom_vline(xintercept = 1,colour="grey40")+
  facet_grid(varname~., switch = "y")+
  labs(title= "Mother and birth characteristics by father participation in cohort") +
    theme_grey(base_family = "Avenir")+
    xlab("Odds ratio\n(odds of the characteristic in pregnancies with a cohort participating father\ncompared to the odds in pregnancies without a participating father)")+
  theme(strip.text.y = element_text(angle = 180),axis.text.y=element_blank(),axis.ticks.y=element_blank(),axis.title.y=element_blank(),panel.grid.minor = element_blank(),panel.grid.major = element_blank(),panel.spacing = unit(.2, "lines"),legend.title=element_blank(),legend.position="bottom",plot.title = element_text(hjust = 0.5))+
   scale_x_continuous(breaks = c(0.5,1,2,3,4) ) +
  coord_trans(x = 'log10')

  
print("alspac")
kable(df_alspac[,1:7])
print("bib")
kable(df_bib[,1:7])
print("mcs")
kable(df_mcs[,1:7])


```


When looking at well-characterised maternal effects (maternal smoking on low birthweight and low gestational age), we see relatively stable effect estimates regardless of whether we use the full sample, the sample where fathers participate or the sample where fathers do not participate (all models adjusted for ethnicity only. The only exception is in ALSPAC and BiB for gestational age: restricting the sample to those with participating fathers (as happens when we adjust for paternal smoking) attenuates the estimate of effect of maternal smoking on gestational age in BiB, whereas in ALSPAC the estimate moves slightly away from the null and instead attenuates if the sample is restricted to non-participating fathers.

```{r , echo=FALSE,include=TRUE,warning=FALSE,message=FALSE}

library(ggplot2)
library(dplyr)
library(ggstance)
library(grid)
library(gridExtra)
library(ggthemr)

ggthemr("light")

## UNADJUSTED FOR ETHNICITY ##

## BiB SA
unadj_ga_bib_sa_res <- rbind.data.frame(c(summary(lm(scale(covs_gestage)~smoking_mother_ever_pregnancy_binary,data=bib_sa_unique))$coef[2,],"Maternal smoking","Complete sample","BiB South Asian","Gestational age","Unadjusted for ethnicity"),
      							c(summary(lm(scale(covs_gestage)~smoking_mother_ever_pregnancy_binary,data=bib_sa_unique[bib_sa_unique$paternal_participation==1,]))$coef[2,],"Maternal smoking","Sample with participating fathers","BiB South Asian","Gestational age","Unadjusted for ethnicity"),
      							c(summary(lm(scale(covs_gestage)~smoking_mother_ever_pregnancy_binary,data=bib_sa_unique[bib_sa_unique$paternal_participation==0,]))$coef[2,],"Maternal smoking","Sample with non-participating fathers","BiB South Asian","Gestational age","Unadjusted for ethnicity"),
      							c(summary(lm(scale(covs_gestage)~smoking_mother_ever_pregnancy_binary+smoking_father_ever_pregnancy_binary,data=bib_sa_unique))$coef[2,],"Maternal smoking","Complete sample, adjusted for paternal smoking","BiB South Asian","Gestational age","Unadjusted for ethnicity"))
colnames(unadj_ga_bib_sa_res)<-c("estimate","se","t","p","parent","model","cohort","outcome","adjusted")

unadj_bw_bib_sa_res <- rbind.data.frame(c(summary(lm(scale(anthro_birthweight)~smoking_mother_ever_pregnancy_binary,data=bib_sa_unique))$coef[2,],"Maternal smoking","Complete sample","BiB South Asian","Birthweight","Unadjusted for ethnicity"),
      							c(summary(lm(scale(anthro_birthweight)~smoking_mother_ever_pregnancy_binary,data=bib_sa_unique[bib_sa_unique$paternal_participation==1,]))$coef[2,],"Maternal smoking","Sample with participating fathers","BiB South Asian","Birthweight","Unadjusted for ethnicity"),
      							c(summary(lm(scale(anthro_birthweight)~smoking_mother_ever_pregnancy_binary,data=bib_sa_unique[bib_sa_unique$paternal_participation==0,]))$coef[2,],"Maternal smoking","Sample with non-participating fathers","BiB South Asian","Birthweight","Unadjusted for ethnicity"),
      							c(summary(lm(scale(anthro_birthweight)~smoking_mother_ever_pregnancy_binary+smoking_father_ever_pregnancy_binary,data=bib_sa_unique))$coef[2,],"Maternal smoking","Complete sample, adjusted for paternal smoking","BiB South Asian","Birthweight","Unadjusted for ethnicity"))
colnames(unadj_bw_bib_sa_res)<-c("estimate","se","t","p","parent","model","cohort","outcome","adjusted")

## BiB WB
unadj_ga_bib_wb_res <- rbind.data.frame(c(summary(lm(scale(covs_gestage)~smoking_mother_ever_pregnancy_binary,data=bib_wb_unique))$coef[2,],"Maternal smoking","Complete sample","BiB White British","Gestational age","Unadjusted for ethnicity"),
      							c(summary(lm(scale(covs_gestage)~smoking_mother_ever_pregnancy_binary,data=bib_wb_unique[bib_wb_unique$paternal_participation==1,]))$coef[2,],"Maternal smoking","Sample with participating fathers","BiB White British","Gestational age","Unadjusted for ethnicity"),
      							c(summary(lm(scale(covs_gestage)~smoking_mother_ever_pregnancy_binary,data=bib_wb_unique[bib_wb_unique$paternal_participation==0,]))$coef[2,],"Maternal smoking","Sample with non-participating fathers","BiB White British","Gestational age","Unadjusted for ethnicity"),
      							c(summary(lm(scale(covs_gestage)~smoking_mother_ever_pregnancy_binary+smoking_father_ever_pregnancy_binary,data=bib_wb_unique))$coef[2,],"Maternal smoking","Complete sample, adjusted for paternal smoking","BiB White British","Gestational age","Unadjusted for ethnicity"))
colnames(unadj_ga_bib_wb_res)<-c("estimate","se","t","p","parent","model","cohort","outcome","adjusted")

unadj_bw_bib_wb_res <- rbind.data.frame(c(summary(lm(scale(anthro_birthweight)~smoking_mother_ever_pregnancy_binary,data=bib_wb_unique))$coef[2,],"Maternal smoking","Complete sample","BiB White British","Birthweight","Unadjusted for ethnicity"),
      							c(summary(lm(scale(anthro_birthweight)~smoking_mother_ever_pregnancy_binary,data=bib_wb_unique[bib_wb_unique$paternal_participation==1,]))$coef[2,],"Maternal smoking","Sample with participating fathers","BiB White British","Birthweight","Unadjusted for ethnicity"),
      							c(summary(lm(scale(anthro_birthweight)~smoking_mother_ever_pregnancy_binary,data=bib_wb_unique[bib_wb_unique$paternal_participation==0,]))$coef[2,],"Maternal smoking","Sample with non-participating fathers","BiB White British","Birthweight","Unadjusted for ethnicity"),
      							c(summary(lm(scale(anthro_birthweight)~smoking_mother_ever_pregnancy_binary+smoking_father_ever_pregnancy_binary,data=bib_wb_unique))$coef[2,],"Maternal smoking","Complete sample, adjusted for paternal smoking","BiB White British","Birthweight","Unadjusted for ethnicity"))
colnames(unadj_bw_bib_wb_res)<-c("estimate","se","t","p","parent","model","cohort","outcome","adjusted")


## ALSPAC
unadj_ga_alspac_res <- rbind.data.frame(c(summary(lm(scale(covs_gestage)~smoking_mother_ever_pregnancy_binary,data=alspac_unique))$coef[2,],"Maternal smoking","Complete sample","ALSPAC","Gestational age","Unadjusted for ethnicity"),
      							c(summary(lm(scale(covs_gestage)~smoking_mother_ever_pregnancy_binary,data=alspac_unique[alspac_unique$paternal_participation==1,]))$coef[2,],"Maternal smoking","Sample with participating fathers","ALSPAC","Gestational age","Unadjusted for ethnicity"),
      							c(summary(lm(scale(covs_gestage)~smoking_mother_ever_pregnancy_binary,data=alspac_unique[alspac_unique$paternal_participation==0,]))$coef[2,],"Maternal smoking","Sample with non-participating fathers","ALSPAC","Gestational age","Unadjusted for ethnicity"),
      							c(summary(lm(scale(covs_gestage)~smoking_mother_ever_pregnancy_binary+smoking_father_ever_pregnancy_binary,data=alspac_unique))$coef[2,],"Maternal smoking","Complete sample, adjusted for paternal smoking","ALSPAC","Gestational age","Unadjusted for ethnicity"))
colnames(unadj_ga_alspac_res)<-c("estimate","se","t","p","parent","model","cohort","outcome","adjusted")

unadj_bw_alspac_res <- rbind.data.frame(c(summary(lm(scale(anthro_birthweight)~smoking_mother_ever_pregnancy_binary,data=alspac_unique))$coef[2,],"Maternal smoking","Complete sample","ALSPAC","Birthweight","Unadjusted for ethnicity"),
      							c(summary(lm(scale(anthro_birthweight)~smoking_mother_ever_pregnancy_binary,data=alspac_unique[alspac_unique$paternal_participation==1,]))$coef[2,],"Maternal smoking","Sample with participating fathers","ALSPAC","Birthweight","Unadjusted for ethnicity"),
      							c(summary(lm(scale(anthro_birthweight)~smoking_mother_ever_pregnancy_binary,data=alspac_unique[alspac_unique$paternal_participation==0,]))$coef[2,],"Maternal smoking","Sample with non-participating fathers","ALSPAC","Birthweight","Unadjusted for ethnicity"),
      							c(summary(lm(scale(anthro_birthweight)~smoking_mother_ever_pregnancy_binary+smoking_father_ever_pregnancy_binary,data=alspac_unique))$coef[2,],"Maternal smoking","Complete sample, adjusted for paternal smoking","ALSPAC","Birthweight","Unadjusted for ethnicity"))
colnames(unadj_bw_alspac_res)<-c("estimate","se","t","p","parent","model","cohort","outcome","adjusted")

## MCS
unadj_ga_mcs_res <- rbind.data.frame(c(summary(lm(scale(covs_gestage)~smoking_mother_ever_pregnancy_binary,data=mcs_unique))$coef[2,],"Maternal smoking","Complete sample","MCS","Gestational age","Unadjusted for ethnicity"),
      							c(summary(lm(scale(covs_gestage)~smoking_mother_ever_pregnancy_binary,data=mcs_unique[mcs_unique$paternal_participation==1,]))$coef[2,],"Maternal smoking","Sample with participating fathers","MCS","Gestational age","Unadjusted for ethnicity"),
      							c(summary(lm(scale(covs_gestage)~smoking_mother_ever_pregnancy_binary,data=mcs_unique[mcs_unique$paternal_participation==0,]))$coef[2,],"Maternal smoking","Sample with non-participating fathers","MCS","Gestational age","Unadjusted for ethnicity"),
      							c(summary(lm(scale(covs_gestage)~smoking_mother_ever_pregnancy_binary+smoking_father_ever_pregnancy_binary,data=mcs_unique))$coef[2,],"Maternal smoking","Complete sample, adjusted for paternal smoking","MCS","Gestational age","Unadjusted for ethnicity"))
colnames(unadj_ga_mcs_res)<-c("estimate","se","t","p","parent","model","cohort","outcome","adjusted")

unadj_bw_mcs_res <- rbind.data.frame(c(summary(lm(scale(anthro_birthweight)~smoking_mother_ever_pregnancy_binary,data=mcs_unique))$coef[2,],"Maternal smoking","Complete sample","MCS","Birthweight","Unadjusted for ethnicity"),
      							c(summary(lm(scale(anthro_birthweight)~smoking_mother_ever_pregnancy_binary,data=mcs_unique[mcs_unique$paternal_participation==1,]))$coef[2,],"Maternal smoking","Sample with participating fathers","MCS","Birthweight","Unadjusted for ethnicity"),
      							c(summary(lm(scale(anthro_birthweight)~smoking_mother_ever_pregnancy_binary,data=mcs_unique[mcs_unique$paternal_participation==0,]))$coef[2,],"Maternal smoking","Sample with non-participating fathers","MCS","Birthweight","Unadjusted for ethnicity"),
      							c(summary(lm(scale(anthro_birthweight)~smoking_mother_ever_pregnancy_binary+smoking_father_ever_pregnancy_binary,data=mcs_unique))$coef[2,],"Maternal smoking","Complete sample, adjusted for paternal smoking","MCS","Birthweight","Unadjusted for ethnicity"))
colnames(unadj_bw_mcs_res)<-c("estimate","se","t","p","parent","model","cohort","outcome","adjusted")


all_res <- rbind.data.frame(unadj_ga_alspac_res,unadj_bw_alspac_res,unadj_bw_bib_sa_res,unadj_ga_bib_sa_res,unadj_bw_bib_wb_res,unadj_ga_bib_wb_res,unadj_bw_mcs_res,unadj_ga_mcs_res)


all_res[,c("estimate","se","t","p")]<-apply(all_res[,c("estimate","se","t","p")],2,function(x) as.numeric(as.character(x)))
all_res$model <- as.character(all_res$model)
all_res$model <- factor(all_res$model, ordered=T, levels = rev(unique(all_res$model)))


##############################################
# Creating plots                             #   
##############################################
ggplot(data=all_res, aes(x=estimate, y=model,colour=adjusted)) + 
  geom_errorbarh(aes(xmax = estimate+(se*1.96), xmin = estimate-(se*1.96)), height = .3, position=position_dodgev(.5)) +
  geom_point(size=2, fill="white",position=position_dodgev(.5),shape=15)+
   geom_vline(aes(xintercept = 0), size = .25, linetype = "dashed", colour= "grey40" ) +
  labs( y = "", x = "Difference in SD outcome in offspring of mothers who smoked\nvs those who did not smoke during pregnancy") +
    facet_grid(outcome~cohort) +
  theme_grey(base_family = "Avenir")+
 theme(panel.grid.minor = element_blank(),panel.grid.major = element_blank(),panel.spacing = unit(.7, "lines"),panel.spacing.y = unit(0.1, "lines"),legend.title=element_blank(),legend.position="none",plot.title = element_text(hjust = 0.5)) +
  labs(title= "Estimates of association between maternal smoking\nduring pregnancy and birthweight and gestational age") 

```


**Negative control**

Not sure if this makes sense. Tried to explain the rationale here and got lost so took out of paper:
Conversely, in ALSPAC only (due to data availability), we also explored associations with two “negative control” outcomes that we would not expect to be affected by parental smoking: family home invasion by mice and family home invasion by pigeons. As described previously4, we hypothesised that associations with both outcomes would be confounded to some extent by socioeconomic position (SEP), but potentially in different directions (mouse infestations are more common in old Victorian and Georgian houses in the UK, which tend to be inhabited by more affluent families, whereas pigeon infestations would be more common in lower SEP families). We conducted logistic regression analysis of associations between maternal and paternal smoking in pregnancy and these negative control outcomes, and considered distance of the point estimate from the null as an indication of the degree of bias  
```{r , echo=FALSE,include=TRUE,warning=FALSE,message=FALSE}
ggthemr("light")

## Unadjusted
unadj_res <- rbind.data.frame(c(summary(glm(negcon_lefthand_binary~smoking_mother_ever_pregnancy_binary,data=alspac_unique,family="binomial"))$coef[2,],"Maternal smoking","Complete sample","ALSPAC","Left handed"),
      							c(summary(glm(negcon_lefthand_binary~smoking_father_ever_pregnancy_binary,data=alspac_unique,family="binomial"))$coef[2,],"Paternal smoking","Complete sample","ALSPAC","Left handed"),
      							c(summary(glm(negcon_mice_binary~smoking_mother_ever_pregnancy_binary,data=alspac_unique,family="binomial"))$coef[2,],"Maternal smoking","Complete sample","ALSPAC","Home invaded by mice"),
      							c(summary(glm(negcon_mice_binary~smoking_father_ever_pregnancy_binary,data=alspac_unique,family="binomial"))$coef[2,],"Paternal smoking","Complete sample","ALSPAC","Home invaded by mice"),
      							c(summary(glm(negcon_pigeons_binary~smoking_mother_ever_pregnancy_binary,data=alspac_unique,family="binomial"))$coef[2,],"Maternal smoking","Complete sample","ALSPAC","Home invaded by pigeons"),
      							c(summary(glm(negcon_pigeons_binary~smoking_father_ever_pregnancy_binary,data=alspac_unique,family="binomial"))$coef[2,],"Paternal smoking","Complete sample","ALSPAC","Home invaded by pigeons"))
colnames(unadj_res)<-c("estimate","se","t","p","parent","model","cohort","outcome")

## mutually-adjusted
adj_res <- rbind.data.frame(c(summary(glm(negcon_lefthand_binary~smoking_mother_ever_pregnancy_binary+smoking_father_ever_pregnancy_binary,data=alspac_unique,family="binomial"))$coef[2,],"Maternal smoking adjusted for paternal smoking","Complete sample","ALSPAC","Left handed"),
      							c(summary(glm(negcon_lefthand_binary~smoking_father_ever_pregnancy_binary+smoking_mother_ever_pregnancy_binary,data=alspac_unique,family="binomial"))$coef[2,],"Paternal smoking adjusted for maternal smoking","Complete sample","ALSPAC","Left handed"),
      							c(summary(glm(negcon_mice_binary~smoking_mother_ever_pregnancy_binary+smoking_father_ever_pregnancy_binary,data=alspac_unique,family="binomial"))$coef[2,],"Maternal smoking adjusted for paternal smoking","Complete sample","ALSPAC","Home invaded by mice"),
      							c(summary(glm(negcon_mice_binary~smoking_father_ever_pregnancy_binary+smoking_mother_ever_pregnancy_binary,data=alspac_unique,family="binomial"))$coef[2,],"Paternal smoking adjusted for maternal smoking","Complete sample","ALSPAC","Home invaded by mice"),
      							c(summary(glm(negcon_pigeons_binary~smoking_mother_ever_pregnancy_binary+smoking_father_ever_pregnancy_binary,data=alspac_unique,family="binomial"))$coef[2,],"Maternal smoking adjusted for paternal smoking","Complete sample","ALSPAC","Home invaded by pigeons"),
      							c(summary(glm(negcon_pigeons_binary~smoking_father_ever_pregnancy_binary+smoking_mother_ever_pregnancy_binary,data=alspac_unique,family="binomial"))$coef[2,],"Paternal smoking adjusted for maternal smoking","Complete sample","ALSPAC","Home invaded by pigeons"))
colnames(adj_res)<-c("estimate","se","t","p","parent","model","cohort","outcome")

all_res <-rbind.data.frame(unadj_res,adj_res)
all_res[,c("estimate","se","t","p")]<-apply(all_res[,c("estimate","se","t","p")],2,function(x) as.numeric(as.character(x)))
all_res$or <- exp(all_res$est)
all_res$upper <- exp(all_res$est + (1.96*all_res$se))
all_res$lower <- exp(all_res$est - (1.96*all_res$se))
all_res$model <- as.character(all_res$model)
all_res$model <- factor(all_res$model, ordered=T, levels = rev(unique(all_res$model)))
all_res$parent <- as.character(all_res$parent)
all_res$parent <- factor(all_res$parent, ordered=T, levels = rev(c("Maternal smoking","Maternal smoking adjusted for paternal smoking","Paternal smoking","Paternal smoking adjusted for maternal smoking")))

negcon_plot<-ggplot(data=all_res[all_res$outcome %in% c("Home invaded by mice","Home invaded by pigeons"),], aes(x=or, y=1,colour=parent)) + 
  geom_errorbarh(aes(xmax = upper, xmin = lower), height = .2, position=position_dodgev(.5)) +
  geom_point(size=2, fill="white",position=position_dodgev(.5),aes(shape=parent))+
   geom_vline(aes(xintercept = 1), size = .25, linetype = "dashed", colour= "grey40" ) +
  facet_grid(outcome~.,switch="y")+
  labs( y = "", x = "Odds ratio\n(odds of outcome in offspring of smokers vs non-smokers") +
    scale_shape_manual(values=c(22,15,22,15),guide=guide_legend(reverse=T,ncol=1))+
  scale_color_manual(values=c('#ffb84d','#ffb84d',"#62bba5","#62bba5"),guide=guide_legend(reverse=T,ncol=2))+
  theme_grey(base_family = "Avenir")+
 theme(strip.text.y = element_text(angle = 180),axis.text.y=element_blank(),axis.ticks.y=element_blank(),panel.grid.minor = element_blank(),panel.grid.major = element_blank(),panel.spacing = unit(.7, "lines"),panel.spacing.y = unit(0.1, "lines"),legend.title=element_blank(),legend.position="right",plot.title = element_text(hjust = 0.5)) +
  #scale_x_continuous(breaks = c(0.5,1,2,3,4) ) +
  coord_trans(x = 'log10')+
  labs(title= "Parents' smoking in pregnancy and negative control outcomes") 
negcon_plot
```

**Rationale for suspecting selection bias**

```{r , echo=FALSE,include=TRUE,warning=FALSE,message=FALSE}

library(broom)

# estimate parameters from alspac data

generate_parameters <- function(dat){
proportions <-  data.frame(
prop.dads = sum(dat[,"paternal_participation"]==1,na.rm=T)/sum(dat[,"paternal_participation"]%in%c(0,1),na.rm=T),
prop.pSmoke = sum(dat[,"smoking_father_ever_pregnancy_binary"]==1,na.rm=T)/sum(dat[,"smoking_father_ever_pregnancy_binary"]%in%c(0,1),na.rm=T),
prop.mSmoke = sum(dat[,"smoking_mother_ever_pregnancy_binary"]==1,na.rm=T)/sum(dat[,"smoking_mother_ever_pregnancy_binary"]%in%c(0,1),na.rm=T))

correlations <- cor(dat[,c("smoking_father_ever_pregnancy_binary","smoking_mother_ever_pregnancy_binary","covs_edu_father","covs_edu_mother","covs_bmi_father_zscore","covs_bmi_mother_zscore","anthro_birthweight")],use="pairwise.complete.obs")

regression_results <- data.frame(pSmoke.pSEP.beta = summary(glm(smoking_father_ever_pregnancy_binary ~ covs_edu_father, data=dat, family="binomial"))$coef[2,1:2],
pBMI.pSmoke.beta = summary(lm(covs_bmi_father_zscore ~ smoking_father_ever_pregnancy_binary, data=dat))$coef[2,1:2],
pBMI.pSEP.beta = summary(lm(covs_bmi_father_zscore ~ covs_edu_father, data=dat))$coef[2,1:2],
pBMI.pSmoke_pSEP.beta = summary(lm(covs_bmi_father_zscore ~ smoking_father_ever_pregnancy_binary + covs_edu_father, data=dat))$coef[2,1:2],

oBW.pBMI.beta = summary(lm(scale(anthro_birthweight) ~ covs_bmi_father_zscore, data=dat))$coef[2,1:2],
oBW.pSmoke.beta = summary(lm(scale(anthro_birthweight) ~ smoking_father_ever_pregnancy_binary, data=dat))$coef[2,1:2],
oBW.pSmoke_pBMI.beta = summary(lm(scale(anthro_birthweight) ~ smoking_father_ever_pregnancy_binary + covs_bmi_father_zscore, data=dat))$coef[2,1:2],

mSmoke.mSEP.beta = summary(glm(smoking_mother_ever_pregnancy_binary ~ covs_edu_mother, data=dat, family="binomial"))$coef[2,1:2],
mBMI.mSmoke.beta = summary(lm(covs_bmi_mother_zscore ~ smoking_mother_ever_pregnancy_binary, data=dat))$coef[2,1:2],
mBMI.mSEP.beta = summary(lm(covs_bmi_mother_zscore ~ covs_edu_mother, data=dat))$coef[2,1:2],
mBMI.mSmoke_mSEP.beta = summary(lm(covs_bmi_mother_zscore ~ smoking_mother_ever_pregnancy_binary + covs_edu_mother, data=dat))$coef[2,1:2],

oBW.mBMI.beta = summary(lm(scale(anthro_birthweight) ~ covs_bmi_mother_zscore, data=dat))$coef[2,1:2],
oBW.mSmoke.beta = summary(lm(scale(anthro_birthweight) ~ smoking_mother_ever_pregnancy_binary, data=dat))$coef[2,1:2],
oBW.mSmoke_mBMI.beta = summary(lm(scale(anthro_birthweight) ~ smoking_mother_ever_pregnancy_binary + covs_bmi_mother_zscore, data=dat))$coef[2,1:2],

pParticipation.mSEP.beta = summary(glm(paternal_participation ~ covs_edu_mother, data=dat, family="binomial"))$coef[2,1:2]
)
list(proportions,regression_results,correlations)
}

res <- lapply(list(alspac_unique,bib_sa_unique,bib_wb_unique,mcs_unique),generate_parameters)
names(res) <- c("alspac","bib south asians","bib white british","mcs")
proportions <- bind_rows(lapply(res,"[[",1))
proportions <- as.data.frame(t(proportions))
names(proportions) <- c("alspac","bib south asians","bib white british","mcs")

reg_res <- bind_rows(lapply(res,"[[",2))
reg_res <- as.data.frame(t(reg_res))
colnames(reg_res) <- c("alspac_coef","alspac_se","bib_south_asians_coef","bib_south_asians_se","bib_white_british_coef","bib_white_british_se","mcs_coef","mcs_se")

reg_res$alspac_lci <- reg_res$alspac_coef - (1.96*reg_res$alspac_se)
reg_res$alspac_uci <- reg_res$alspac_coef + (1.96*reg_res$alspac_se)
reg_res$bib_south_asians_lci <- reg_res$bib_south_asians_coef - (1.96*reg_res$bib_south_asians_se)
reg_res$bib_south_asians_uci <- reg_res$bib_south_asians_coef + (1.96*reg_res$bib_south_asians_se)
reg_res$bib_white_british_lci <- reg_res$bib_white_british_coef - (1.96*reg_res$bib_white_british_se)
reg_res$bib_white_british_uci <- reg_res$bib_white_british_coef + (1.96*reg_res$bib_white_british_se)
reg_res$mcs_lci <- reg_res$mcs_coef - (1.96*reg_res$mcs_se)
reg_res$mcs_uci <- reg_res$mcs_coef + (1.96*reg_res$mcs_se)

reg_res <- reg_res[,c("alspac_coef","alspac_se","alspac_lci","alspac_uci",
           "bib_white_british_coef","bib_white_british_se","bib_white_british_lci","bib_white_british_uci",
           "bib_south_asians_coef","bib_south_asians_se","bib_south_asians_lci","bib_south_asians_uci",
           "mcs_coef","mcs_se","mcs_lci","mcs_uci")]

correlations_alspac <- lapply(res,"[[",3)$alspac
correlations_bib_sa <- lapply(res,"[[",3)$"bib south asians"
correlations_bib_wb <- lapply(res,"[[",3)$"bib white british"
correlations_mcs <- lapply(res,"[[",3)$"mcs"

write.csv(reg_res,"regression_parameters_for_louise.csv",row.names=T)
write.csv(proportions,"proportions_for_louise.csv",row.names=T)
write.csv(correlations_alspac,"correlations_alspac_for_louise.csv",row.names=T)
```